---
title: "Diagnostics How-To"
author:
  - Mike Ackerman:
      email: mike.ackerman@merck.com
      institute: [biomark]
      correspondence: true
  - Kevin See:
      email: kevin.see@merck.com
      institute: [biomark]
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      fig_height: 6
      fig_width: 6
      toc: yes
      toc_depth: 3
      toc_float:
        collapsed: yes
        smooth_scroll: yes
      theme: flatly
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
vignette: >
  %\VignetteIndexEntry{Fitting a CJS Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
abstract: |
  This manual contains instructions on how to read in telemetry records and perform diagnostics including evaluations of receiver noise and operations and tag battery life evaluations.
---    

```{r setup, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
# knitr library and options
library(knitr)
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>"
)

# load necessary libraries
library(telemetyr)
library(dplyr)

```

# Introduction

This vignetted describes how to read in various files downloaded during the field season using the Tracker software and then perform various "diagnostics" using data within those files. The diagnostics include summaries of receiver operation times, noise, and voltage/temperature and test tag battery life. Similar diagnostic summaries e.g. of sentinel tag detections could easily be summarised and added to the \code{telemetyr} R package at a later date. 

# Data Prep

The raw data is contained in *.txt* files, located in a folder named after the receiver that it was downloaded from. Each receiver has a 3-digit code. And typically, all of the receiver folders for a given study season are stored in a dirctory for that season. So first, let's set the file path to the folder that contains all of the reciever folders. The following provides an example for where data for the 2018/2019 fixed site download data is stored for the Lemhi River juvenile Chinook telemetry.

```{r data-path}
download_path = "S:/telemetry/lemhi/fixed_site_downloads/2018_2019"
# download_path = "~/../../Volumes/ABS/telemetry/lemhi/fixed_site_downloads/2018_2019"
```

After setting the path to the receiver downloads, use the `read_txt_data()` function to read in each text file and combine them all into one long dataframe. 

```{r read-raw-data, eval = F}
# read in raw .txt data
raw_df = read_txt_data(download_path)

```

You can use the `save()` function to save the above `raw_df` object (or whatever you called it), perhaps as a *.rda* file, to `load()` at a later time. Note, loading the `raw_df` object saves substantial time over running `read_txt_data()` each time. 

```{r load-raw-data, eval = F}
# load raw_df
load("../data/prepped/2018_2019/raw.rda")

```

The next step is to clean up some errant dates in the receiver records. We noted that many files have incorrect dates at the beginning of the file, perhaps because the timer had not been reset after the last download. We build a function `clean_raw_data()`, that fixes as many of those dates as possible, based on the valid dates within each file. In addition, `clean_raw_data()` fixes any file that lists the receiver as `000`, and has the option to filter out records that are marked as invalid (`valid == 0`). The function also includes the arguments `min_yr` and `max_yr` to put bounds around the expected dates, to help determine which dates should be corrected.

After that, we wanted to round the tag code that was recorded, to the nearest possible code possible, hopefully to match a tag that was actually deployed. In some years, tags ending in a 0 or 5 were deployed, in others only tags ending in 0. So we've added another function, `round_tag_codes()`, that includes as an input parameter as to whether to round detected tag codes to the nearest 5 or nearest 10, and rename it to `tag_id`, after which we assume that matches up to a tag ID from a tag actually deployed.

Finally, we compressed those detections or records into windows of time, such that for each tag, we record how many observations occurred within that window. The default time window is 2 minutes, but that can easily be changed by the user. The compression of the data is performed using the function `compress_txt_data()`.

For convenience, all of the above functionality is wrapped into a single wrapper function `compress_raw_data()`. The resulting compressed data appears to match the format of the data provided by the .csv files that are also downloaded from the receivers.

Finally, we compressed these detections into windows of time, such that for each tag, we record how many observations occurred within that window. The default time window is 2 minutes, but that can be changed by the user. All of the above functionality is wrapped into a single wrapper function, `compress_raw_data()`. This compressed data appears to match the format of the data provided by the *.csv* files that are also downloaded from the receivers. 

```{r compress-raw-data, echo = F, eval = F}
# clean, round and compress data
compress_df = compress_raw_data(raw_df,
                                min_yr = 2018,
                                max_yr = 2019,
                                round_to = 5)

```

Again, the `compress_df` object can be saved using `save()` and loaded at a later date, thus saving time. The `compress_raw_data()` function can take a fair amount of time depending on the size of your dataset.

```{r load-data, echo = F}
load('../data/prepped/2018_2019/compressed.rda')

```

The compressed data looks something like this:

```{r view-data}
compress_df %>%
  filter(receiver != 'ACT') %>%
  head()
```

From here on out, unless otherwise notes, we will perform our "diagnostics" using the `compress_df` object which contains our cleaned and rounded data in a compressed format. In the [Receiver Volt and Temperature](#receiver-volt-and-temperature) section, we will have to read in that information, which are stored in separate *.txt* files within the `download_path`.

# Diagnostics

## Receiver Names

Most of the functions in the subsequent sections use an argument `receiver_codes` that can be used to specify which receivers you'd like to evaluate. If `receiver_codes = NULL`, the default, the function will simply be performed on all unique receivers present within the `compress_df`. The following are a few examples of how the user can create an object `receiver_nms`, that can later be fed into the `receiver_codes` argument.

```{r receiver-names}
# a list of the receiver names; note using this option would 
# provide the same output as if receiver_codes = NULL
receiver_nms = unique(compress_df$receiver)

# provide an explicit list of receiver codes
receiver_nms = c("CA1", "CC1", "DC1", "DG1", "DW1", "KP1", "LH1", "LR1", "NF1",
                 "RR1", "SB1", "SR1", "TB1", "TR1", "TT1", "TT2", "VC1", "YP1")

# get all unique receivers, but then discard a few
receiver_nms = unique(compress_df$receiver) %>%
  setdiff(c("ACT", "TT1", "TT2"))

```

## Receiver Operation Times

We've provided a function, `summarise_timer_data()`, that uses records in the data from timer tags, typically `tag_id` ending in "575" to summarise the hours for which each receiver was operational or not. The function should be performed on the compressed data. Optionally, the `receiver_codes` argument can be used to summarise operational times for a subset of receivers. By default, the function uses the first and last observation within the timer tag data for plotting and the first and last observation for each receiver to calculate the proportion of time a given receiver was operational. However, the user can use the `season_start` and `season_end` arguments ("%Y-%m-%d" i.e. "2018-09-15" format) to override those dates, in which case, the x-axis of the plot and the estimated proportion operational times will reflect those dates. The following are a few examples of how to use the `summarise_timer_data()` function to create a `timer_summ` object.

```{r timer, results = "hide"}
# summarise timer data - all receivers in compress_df
timer_summ = summarise_timer_data(compress_df)

# summarise timer data - only receiver_codes in receiver_nms
timer_summ = summarise_timer_data(compress_df,
                                  receiver_codes = receiver_nms)

# summarise timer data - only receiver_codes in receiver_nms, and
# manually set the season start and season end dates
timer_summ = summarise_timer_data(compress_df,
                                  receiver_codes = receiver_nms,
                                  season_start = "2018-09-15",
                                  season_end   = "2019-04-01")
```

The resulting `timer_summ` list contains a few objects summarizing operatinal times. The `timer_summ$operations_summ` object contains a record for each hour of the season showing whether a given receiver was operational (`TRUE`) or not (`FALSE`). The `timer_summ$p_operational` data frame provides as estimate for each receiver of the proportion of time that the receiver was operational during the season. Finally, `timer_summ$operations_plot` provides a plot showing the operational times for each receiver. The following are examples of the objects available from `summarise_timer_data()`:

```{r timer-objects, fig.cap = "A plot of receiver operation times. Hours for which the receiver were operational are shown as black."}
# TRUE/FALSE summary of receiver operations by hour
head(timer_summ$operations_summ)

# proportion of time that each receiver was operational
head(timer_summ$p_operational)

# plot showing receiver operation times
timer_summ$operations_plot       

```

## Receiver Noise

The `summarise_noise_data()` functions summarises "noise" records from each receiver, typically `tag_id` ending in "995". Again, this function should be performed on the compressed data and the `receiver_codes` argument can be used to look at only a subset of receivers. Optionally, the `operations_summary` argument, typically the `operations_summ` object in the `timer_summ` list (see [Receiver Operation Times](#receiver-operation-times) section), can be used to calculate a noise rate (i.e. noise records per hour). If `operations_summary = NULL`, the default, the function will simply tally the number of noise records by receiver and channel.

```{r noise, results = "hide"}
# summarise noise data - all receivers in compress_df
noise_summ = summarise_noise_data(compress_df)

# summarise noise data - only receiver_codes in receiver_nms
noise_summ = summarise_noise_data(compress_df,
                                  receiver_codes = receiver_nms)

# summarise noise data - only receiver_codes in receiver_nms,
# and provide receiver operations summary so that noise can 
# be converted to a rate (noise/hr)
noise_summ = summarise_noise_data(compress_df,
                                  receiver_codes = receiver_nms,
                                  operations_summary = timer_summ$operations_summ)

```

The resulting `noise_summ` list contains two objects. The `noise_tbl` object contains a simple summary of noise by receiver and channel and the `noise_plot` provides a heat map of that table. The following are examples of those objects available from the 2018/2019 season in the Lemhi River dataset:

```{r noise-objects, fig.cap = "A heat map summarising noise records by receiver and channel."}
# summary of noise by receiver and channel
head(noise_summ$noise_tbl) 

# a plot of noise_tbl
noise_summ$noise_plot 

```

## Test Tag Battery Life

The `summarise_test_data()` function is provided to summarise data available from the "test" tags, typically to estimate the battery life for tags, and this function should also be run on the compressed data. A path to the metadata for all tags deployed, including test tags, should be provided to the argument `tag_data` and that .xlsx file should contain two columns `radio_tag_id` and `duty_cycle` containing the tag code and batch information for each tag, respectively. By default, the function will also use dates in the `compress_df` object to determine the appropriate season and only grab test tags for that season.

NOTE: The `summarise_test_data()` function could easily be modified by the authors, in the future, so that the user could provide their own list of test tags and duty cycles.

```{r test, results = "hide"}
# summarise test tag data
test_summ = summarise_test_data(compress_df,
                                tag_data = "../data/prepped/tag_release/lemhi_winter_telemetry_tag_info.xlsx")

```

The `summarise_test_data()` function, like the other summary functions, also returns a list containing a handful of objects. The `test_tag_ids` is simply the test tag IDs and duty cycle information extracted for a given season. The `test_df` object is a data frame containing the activation time and time that each tag was determined to be "dead" along with an estimate tag life using all of the records for those test tags. The duty cycle for each tag is also provided for convenience. The `tag_life` provides a quick summary of tag life including the mean standard, deviation, and quartiles, and finally, the `tag_life_p` is a boxplot of battery life information, summarised by duty cycle. The following are examples of those objects from the `summarise_test_data()` function.

```{r test-objects, fig.cap = "test."}
# list of test tags and their duty cycle
head(test_summ$test_tag_ids)

# a summary of tag life for each test tag
head(test_summ$test_df)

# summary by duty cycle
test_summ$tag_life

# a box plot of tag life by duty cycle
test_summ$tag_life_p   

```

## Receiver Volt and Temperature

Finally, a user may be interested in examining or plotting receiver voltage or temperature data from the receivers. This information is stored in different files within `download_path`, typically ending in *\$\$.txt*, separate from the observation data and not in the `compress_df`. We have provided a function `read_volt_temp_data()` to read in and combine all of those files ending in *\$\$.txt*.

```{r volt-temp, results = "hide"}
# read receiver volt & temperature info
vt_df = read_volt_temp_data(download_path)

```

The user can then view or work with the `vt_df` data frame as they see fit. For quick plotting, we have provided a function `plot_volt_temp_data()` to plot a given variable within `vt_df`, faceted by receiver. The user can use the `column` argument to specify which column in `vt_df` they would like to plot, and again, the `receiver_codes` function can be used to plot only a subset of receivers.

```{r volt-temp-plot, fig.cap = "A plot of voltage, faceted by receiver."}
# plot voltage information
volt_p = plot_volt_temp_data(vt_df,
                             column = "volt_avg")

# plot voltage information - only receiver_codes in receiver_nms
volt_p = plot_volt_temp_data(vt_df,
                             column = "volt_avg",
                             receiver_codes = receiver_nms)

# plot temperature information - only receiver_codes in receiver_nms
temp_p = plot_volt_temp_data(vt_df,
                             column = "temp_avg",
                             receiver_codes = receiver_nms)

# view the last volt_p
volt_p

```

#### END DIAGNOSTICS VIGNETTE
