---
title: "PIT Tag-Only CJS Models"
author:
  - Mike Ackerman:
      email: mike.ackerman@merck.com
      institute: [biomark]
      correspondence: true
  - Kevin See:
      email: kevin.see@merck.com
      institute: [biomark]
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::html_vignette2:
    fig_height: 6
    fig_width: 6
    pandoc_args:
    - --lua-filter=templates/scholarly-metadata.lua
    - --lua-filter=templates/author-info-blocks.lua
  rmarkdown::html_vignette:
    fig_height: 6
    fig_width: 6
    pandoc_args:
    - --lua-filter=templates/scholarly-metadata.lua
    - --lua-filter=templates/author-info-blocks.lua
  bookdown::html_document2:
    fig_height: 6
    fig_width: 6
    theme: flatly
    pandoc_args:
    - --lua-filter=templates/scholarly-metadata.lua
    - --lua-filter=templates/author-info-blocks.lua
vignette: >
  %\VignetteIndexEntry{Fitting a CJS Model}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = F,
  message = F,
  comment = "#>"
)
```

```{r setup, message=F}
# setwd('vignettes')
library(telemetyr)
library(tidyverse)
library(readxl)
library(lubridate)
library(magrittr)
library(PITcleanr) # to install, use devtools::install_github("KevinSee/PITcleanr")
library(janitor)
library(rjags)
library(postpack)
# for formatting in this vignette
library(kableExtra)

theme_set(theme_bw())
```

# Introduction

We are interested primarily in estimating survival of fish from release (e.g. at the Lower Lemhi screw trap) to arrival at Lower Granite dam, using a Cormack Jolly-Seber (CJS) model. 

# Methods

## Data

The data is PIT tag detections, which we queried from [PTAGIS](www.ptagis.org). For a given list of tags, we queried their complete detection history, including fields such as:

* Tag Code
* Mark Rear Type
* Event Type
* Event Date/Time
* Event Release Date/Time
* Antenna ID
* Antenna Group Configuration Value

We then consolidated some of the detections, to simplify the model. We were primarily focused on survival between the lower Lemhi array (LLR) and Lower Granite Dam (GRJ). All the sites with a river kilometer downstream of Lower Granite were grouped into a "Below_GRJ" site, and any intermediate detections at sites between LLR and GRJ were grouped in with LLR detections, since we know that fish made it past LLR. This leaves us with a matrix of 4 columns: release, LLR, GRJ and Below_GRJ, with one row per tag and a 1 or 0 in the column corresponding to whether that tag was detected at each of those sites. 

## Model

We used a Cormack Jolly-Seber model to estimate survival between reaches and detection probability at each detection point. The detection probability at the last site and the survival to that site are confounded in the model, and cannot be estimated separately. This is why we ensured our last site was below Lower Granite dam. 

```{r prep-data}
# read in all the information about radio tagged fish
rt_fish_df = list('pilot',
                  '2018_2019',
                  '2019_2020') %>%
  rlang::set_names() %>%
  map_df(.id = 'Year',
         .f = function(x) {
           load(paste0('../data/prepped/', x, '/cap_hist.rda'))
           return(cap_hist_list$tag_df)
         }) %>%
  # filter out fish with no PIT tag information
  filter(!is.na(pit_tag_id))

# build configuration table (requires internet connection and PITcleanr package)
org_config = buildConfig()

# read in detections from PTAGIS
observations = read_csv('../data/raw/PITcleanr_query.csv')

# compress observations
comp_df = observations %>%
  clean_names(case = 'snake') %>%
  mutate_at(vars(matches('date_time')),
            list(mdy_hms)) %>%
  group_by(tag_code) %>%
  mutate(season = year(event_date_time_value[event_type_name == "Mark"])) %>%
  ungroup() %>%
  mutate(obs_date = if_else(is.na(event_date_time_value),
                            event_release_date_time_value,
                            event_date_time_value)) %>%
  group_by(season, tag_code, site_code = event_site_code_value) %>%
  summarise(min_date = min(obs_date),
            max_date = max(obs_date),
            n_obs = n()) %>%
  ungroup()

# consolidate some sites in the configuration file
configuration = org_config %>%
  filter(SiteID %in% unique(comp_df$site_code)) %>%
  filter(!EndDate < min(comp_df$min_date) | is.na(EndDate)) %>%
  mutate(Node = SiteID) %>%
  mutate(Node = if_else(RKMTotal < unique(RKMTotal[SiteID == "GRJ"]),
                        'Below_GRJ',
                        Node),
         Node = if_else(RKMTotal < unique(RKMTotal[SiteID == "LLR"]) &
                          RKMTotal > unique(RKMTotal[SiteID == "GRJ"]) &
                          SiteID != "LEMHIR",
                        "LLR",
                        Node),
         Node = if_else(SiteID == "LEMHIR",
                        "LLRTP",
                        Node)) %>%
  select(site_code = SiteID, Node, SiteType, SiteName, RKM, RKMTotal) %>%
  distinct() %>%
  arrange(desc(RKMTotal))

# build capture histories
cap_hist = comp_df %>%
  left_join(configuration %>%
              group_by(Node) %>%
              mutate(min_rkm = min(RKMTotal))) %>%
  mutate(Node = fct_reorder(Node,
                            min_rkm),
         Node = fct_rev(Node)) %>%
  group_by(season, tag_code, Node) %>%
  summarise(min_date = min(min_date),
            max_date = max(max_date),
            n_obs = sum(n_obs)) %>%
  ungroup() %>%
  select(-ends_with("date")) %>%
  mutate(n_obs = if_else(n_obs > 0, as.integer(1), n_obs)) %>%
  spread(Node, n_obs,
         fill = 0) %>%
  # mutate(Release = 1) %>%
  select(season, tag_code,
         LLRTP:Below_GRJ)

cap_hist %>%
  left_join(rt_fish_df %>%
              select(tag_code = pit_tag_id,
                     release_site)) %>%
  filter(release_site == "LLRTP") %>%
  # filter(season == 2018) %>%
  filter(LLRTP == 0,
         LLR == 0,
         GRJ == 0,
         Below_GRJ == 0)


# develop list of input matrices for CJS model
# focus on fish released from lower Lemhi screw trap or detected there
y_list = cap_hist %>%
    left_join(rt_fish_df %>%
              select(tag_code = pit_tag_id,
                     release_site)) %>%
  filter(release_site == "LLRTP" |
           LLRTP == 1) %>%
  select(-release_site) %>%
  split(list(.$season)) %>%
  map(.f = function(x) {
    x %>%
      select(-(season:tag_code)) %>%
      as.matrix()
  })

```

```{r jags-model}
# specify model in JAGS
jags_model = function() {
  # PRIORS
  phi[1] <- 1
  p[1] <- 1
  for(j in 2:J) {
    phi[j] ~ dbeta(1,1) # survival probability between arrays
    p[j] ~ dbeta(1,1)   # detection probability at each array
  }

  # LIKELIHOOD - Here, p and phi are global
  for (i in 1:N) {
    # first known occasion must be z == 1
    # z[i, f[i]] <- 1
    # j = 1 is the release occasion - known alive; i.e., the mark event
    for (j in (f[i] + 1):J) {
      # survival process: must have been alive in j-1 to have non-zero pr(alive at j)
      z[i,j] ~ dbern(phi[j] * z[i,j-1]) # fish i in period j is a bernoulli trial

      # detection process: must have been alive in j to observe in j
      y[i,j] ~ dbern(p[j] * z[i,j]) # another bernoulli trial
    }
  }

  # DERIVED QUANTITIES
  # survivorship is probability of surviving from release to a detection occasion
  survship[1] <- 1 # the mark event; everybody survived to this point
  for (j in 2:J) { # the rest of the events
    survship[j] <- survship[j-1] * phi[j]
  }
}

# write model to a text file
jags_file = "CJS_model.txt"
write_model(jags_model, jags_file)

# specify which parameters to track
jags_params = c("phi", "p", "survship")
```

```{r fit-jags-model, eval = F}
# only focus on 2017-18 and 2018-19 for now, since PIT tag detections at Lower Granite and downstream may not be recorded until after June 30
for(i in 1:2) {
  
  y = y_list[[i]]
  
  # put together data for JAGS
  jags_data = list(
    N = nrow(y),
    J = ncol(y),
    y = y,
    z = known_alive(y),
    f = first_alive(y)
  )
  
  # using rjags package
  jags = jags.model(jags_file,
                    data = jags_data,
                    n.chains = 4,
                    n.adapt = 1000)
  # burnin
  update(jags, n.iter = 2500)
  # posterior sampling
  post = coda.samples(jags,
                      jags_params,
                      n.iter = 2500,
                      thin = 5)
  
  # posterior summaries
  param_summ = post_summ(post,
                         jags_params,
                         Rhat = T,
                         ess = T) %>%
    t() %>%
    as_tibble(rownames = "param") %>%
    mutate(cv = sd / mean)
  
  # save model results
  file_nm = paste0('PIT_only_CJS_', names(y_list)[i], '.rda')
  
  save(jags_data, post, param_summ,
       file = paste0("../analysis/CJS_models/", file_nm))
  
  rm(y, jags_data, jags, post, param_summ, file_nm)
}

```


# Results

This CJS model has 4 detection points: lower Lemhi trap, LLR, GRJ and anywhere downstream of GRJ. Survival and detection probabilities for the last detection point are confounded in CJS models, so we do not present those estimates. Detection at the release site (lower Lemhi trap) is fixed at 100%, since that's the most upstream site and where fish enter the model. Therefore, we focus on survival between the trap and LLR, and between LLR and GRJ, (Table \ref{tab:phi-tab}) as well as detection probabilities at LLR and GRJ (Table \ref{tab:det-tab}). In addition, we calculated cummulative survival to each site as well (Table \ref{tab:surv-tab}). 

```{r}
# compile parameter estimaets for all years
param_summ = 2017:2018 %>%
  as.list() %>%
  rlang::set_names() %>%
  map_df(.id = 'season',
         .f = function(yr) {
           load(paste0("../analysis/CJS_models/PIT_only_CJS_", yr, ".rda"))
           return(param_summ)
         }) %>%
  mutate(season = recode(as.character(season),
                         "2017" = "2017-18",
                         "2018" = "2018-19",
                         "2019" = "2019-20")) %>%
  # filter out parameters that are either fixed at 1, or not actually estimable
  filter(!grepl('\\[1\\]$', param),
         !grepl('\\[4\\]$', param))


# labels for plotting
site_nms = c("LLR",
             "GRJ")
between_nms = c("LLRTP to LLR",
                "LLR to GRJ")


```

```{r phi-tab}
param_summ %>%
  filter(grepl('phi', param)) %>%
  mutate(site = factor(rep(between_nms, n_distinct(season)),
                       levels = between_nms)) %>%
  select(season, Reach = site, 
         mean, 
         se = sd,
         cv,
         lwrCI = `2.5%`,
         uprCI = `97.5%`) %>%
  kable(caption = 'Estimates of survival between detection points.',
        digits = 3) %>%
  kable_styling()
```


```{r det-tab}
param_summ %>%
  filter(grepl('^p\\[', param)) %>%
  mutate(site = factor(rep(site_nms, n_distinct(season)),
                       levels = site_nms)) %>%
  select(season, site, 
         mean, 
         se = sd,
         cv,
         lwrCI = `2.5%`,
         uprCI = `97.5%`) %>%
  kable(caption = 'Estimates of detection probability.',
        digits = 3) %>%
  kable_styling()
```


```{r surv-tab}
param_summ %>%
  filter(grepl('surv', param)) %>%
  mutate(site = factor(rep(site_nms, n_distinct(season)),
                       levels = site_nms)) %>%
  select(season, site, 
         mean, 
         se = sd,
         cv,
         lwrCI = `2.5%`,
         uprCI = `97.5%`) %>%
  kable(caption = 'Estimates of cummulative survival up to detection points.',
        digits = 3) %>%
  kable_styling()
```

We also present these results in Figures \ref{fig:phi-fig}, \ref{fig:det-fig} and \ref{fig:surv_fig}.

```{r}
# plots of estimates
dodge_width = 0.3


surv_p = param_summ %>%
  filter(grepl('survship', param)) %>%
  mutate(site = factor(rep(site_nms, n_distinct(season)),
                       levels = site_nms)) %>%
  ggplot(aes(x = site,
             y = mean,
             color = season)) +
  scale_color_brewer(palette = "Set1") +
  geom_errorbar(aes(ymin = `2.5%`,
                    ymax = `97.5%`),
                width = 0,
                position = position_dodge(width = dodge_width)) +
  geom_point(position = position_dodge(width = dodge_width)) +
  labs(x = 'Site',
       y = 'Cummulative Survival',
       color = 'Season')

phi_p = param_summ %>%
  filter(grepl('phi', param)) %>%
  mutate(site = factor(rep(between_nms, n_distinct(season)),
                       levels = between_nms)) %>%
  ggplot(aes(x = site,
             y = mean,
             color = season)) +
  scale_color_brewer(palette = "Set1") +
  geom_errorbar(aes(ymin = `2.5%`,
                    ymax = `97.5%`),
                width = 0,
                position = position_dodge(width = dodge_width)) +
  geom_point(position = position_dodge(width = dodge_width)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = 'Reach',
       y = 'Survival Within Reach',
       color = 'Season')

det_p = param_summ %>%
  filter(grepl('^p\\[', param)) %>%
  mutate(site = factor(rep(site_nms, n_distinct(season)),
                       levels = site_nms)) %>%
  ggplot(aes(x = site,
             y = mean,
             color = season)) +
  scale_color_brewer(palette = "Set1") +
  geom_errorbar(aes(ymin = `2.5%`,
                    ymax = `97.5%`),
                width = 0,
                position = position_dodge(width = dodge_width)) +
  geom_point(position = position_dodge(width = dodge_width)) +
  labs(x = 'Site',
       y = 'Detection Probability',
       color = 'Season')

```

```{r phi-fig, fig.cap = "Estimates, with 95% confidence intervals, of survival between reaches."}
phi_p
```

```{r det-fig, fig.cap = "Estimates, with 95% confidence intervals, of detection probabilities."}
det_p
```

```{r surv-fig, fig.cap = "Estimates, with 95% confidence intervals, of cummulative survival between reaches."}
surv_p
```
