---
title: "Radio Tag-Only CJS Models"
author:
  - Kevin See:
      email: kevin.see@merck.com
      institute: [biomark]
      correspondence: true
  - Mike Ackerman:
      email: mike.ackerman@merck.com
      institute: [biomark]
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::html_document2:
    fig_height: 6
    fig_width: 8
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    theme: flatly
    pandoc_args:
    - --lua-filter=templates/scholarly-metadata.lua
    - --lua-filter=templates/author-info-blocks.lua
vignette: >
  %\VignetteIndexEntry{Fitting a CJS Model}
  %\VignetteEncoding{UTF-8}
  <!-- %\VignetteEngine{knitr::rmarkdown} -->
  %\VignetteEngine{bookdown::bookdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = F,
  message = F,
  echo = F,
  comment = "#>"
)
```

```{r setup, message=F}
# setwd('analysis/writeups')
library(telemetyr)
library(tidyverse)
library(readxl)
library(lubridate)
library(magrittr)
library(janitor)
library(rjags)
library(postpack)
# for formatting in this vignette
library(kableExtra)

theme_set(theme_bw())
```

# Introduction

We are interested primarily in estimating survival of fish from release (e.g. at the Lower Lemhi screw trap) between each RT site using radio tag detections only, with a Cormack Jolly-Seber (CJS) model.

# Methods

## Data

The data is radio tag (RT) detections, which were downloaded from various fixed sites over the last three years. We compressed these detections and summarised capture histories, eliminating any detections that indicated upstream movement. 

```{r prep-data}
#-------------------------
# read in RT site metadata
#-------------------------
rec_meta = read_excel('../../data/prepped/site_metadata/rt_site_metadata.xlsx')

# which RT receivers were used each year?
rec_df = rec_meta %>%
  filter(site_type == 'rt_fixed') %>%
  filter(as.integer(str_sub(rt_rkm, end = 3)) <= 274) %>%
  mutate_at(vars(site_code),
            list(~ factor(., levels = unique(.)))) %>%
  gather(season, use, starts_with("use")) %>%
  mutate(season = str_remove(season, "use")) %>%
  filter(use) %>%
  select(-use) %>%
  select(season, everything())

#-------------------------
# load capture histories for all years
#-------------------------
# load all years of data
rt_ch = list('pilot',
     '2018_2019',
     '2019_2020') %>%
  rlang::set_names() %>%
  map_df(.id = 'Year',
         .f = function(x) {
           load(paste0('../../data/prepped/', x, '/cap_hist.rda'))
           ch = cap_hist_list$tag_df %>%
             left_join(cap_hist_list$ch_wide) %>%
             mutate(some_det = if_else(is.na(cap_hist), F, T)) %>%
             filter(duty_cycle == 'batch_1' | some_det) %>%
             select(-some_det)

           return(ch)
         }) %>%
  select(-cap_hist) %>%
  # filter out releases from PAHTRP and HYDTRP
  filter(release_site %in% c("LEMTRP", "LLRTP")) %>%
  # filter out on-off duty cycles
  filter(duty_cycle != "on_off") %>%
  mutate(LLRTP = if_else(release_site == 'LLRTP',
                         1, 0),
         LEMTRP = if_else(release_site == 'LEMTRP',
                         1, 0)) %>%
  # put sites in correct order for CJS model
  gather(site, seen, -(Year:tag_id)) %>%
  mutate(site = factor(site,
                       levels = c("LEMTRP", levels(rec_df$site_code), "LLRTP")),
         site = fct_relevel(site, "LLRTP", after = 8)) %>%
  filter(!is.na(site)) %>%
  spread(site, seen,
         fill = 0)


# develop list of input matrices for CJS model
# drop columns from sites that were not operating in that years
y_list = rt_ch %>%
  split(list(.$season)) %>%
  map(.f = function(x) {
    if(x$season %in% c('17_18', '18_19')) {
      sites = rec_df %>%
        filter(season == unique(x$season)) %>%
        pull(site_code) %>%
        as.character()
      y = x %>%
        select(one_of(c('LLRTP', sites))) %>%
        select(one_of(names(x))) %>%
        as.matrix()
    }

    if(x$season == '19_20') {
      sites = rec_df %>%
        filter(season == unique(x$season)) %>%
        pull(site_code) %>%
        as.character()
      y = x %>%
        # filter out fish released from lower Lemhi trap but detected upstream
        filter(!(release_site == 'LLRTP' &
                 (BC == 1 | TC == 1 | EC == 1 | SS == 1 | EU == 1 | LF == 1))) %>%
        select(one_of(c("LEMTRP", 'LLRTP', sites))) %>%
        select(one_of(names(x))) %>%
        as.matrix()
    }
    return(y)
  })

```

In 2017-18 and 2018-19, we focused on tags that were released from the lower Lemhi rotary screw trap (code LLRTP). In 2019-20, we also included tags released from the upper Lemhi rotary screw trap (code LEMTRP). We did remove any tags released at the lower Lemhi trap that were detected upstream in the Lemhi River. There were `r filter(rt_ch, (release_site == 'LLRTP' & (BC == 1 | TC == 1 | EC == 1 | SS == 1 | EU == 1 | LF == 1))) %>% nrow` tags with a detection history like this.

## Model

We used a Cormack Jolly-Seber model to estimate survival between reaches and detection probability at each detection point. The detection probability at the last site and the survival to that site are confounded in the model, and cannot be estimated separately, and so results from the final reach each year are excluded.


```{r jags-model, eval = F}
# specify model in JAGS
jags_model = function() {
  # PRIORS
  phi[1] <- 1
  p[1] <- 1
  for(j in 2:J) {
    phi[j] ~ dbeta(1,1) # survival probability between arrays
    p[j] ~ dbeta(1,1)   # detection probability at each array
  }

  # LIKELIHOOD - Here, p and phi are global
  for (i in 1:N) {
    # first known occasion must be z == 1
    # z[i, f[i]] <- 1
    # j = 1 is the release occasion - known alive; i.e., the mark event
    for (j in (f[i] + 1):J) {
      # survival process: must have been alive in j-1 to have non-zero pr(alive at j)
      z[i,j] ~ dbern(phi[j] * z[i,j-1]) # fish i in period j is a bernoulli trial

      # detection process: must have been alive in j to observe in j
      y[i,j] ~ dbern(p[j] * z[i,j]) # another bernoulli trial
    }
  }

  # DERIVED QUANTITIES
  # survivorship is probability of surviving from release to a detection occasion
  survship[1] <- 1 # the mark event; everybody survived to this point
  for (j in 2:J) { # the rest of the events
    survship[j] <- survship[j-1] * phi[j]
  }
}

# write model to a text file
jags_file = "../CJS_models/CJS_model.txt"
write_model(jags_model, jags_file)

# specify which parameters to track
jags_params = c("phi", 
                "p", 
                "survship")

```

```{r fit-jags-model, eval = F}
for(i in 1:3) {
  cat(paste("Working on", names(y_list)[i], "\n"))

  y = y_list[[i]]

  jags_data = list(
    N = nrow(y),
    J = ncol(y),
    y = y,
    z = known_alive(y),
    f = first_alive(y)
  )

  # using rjags package
  jags = jags.model(jags_file,
                    data = jags_data,
                    n.chains = 4,
                    n.adapt = 1000)
  # burnin
  update(jags, n.iter = 2500)
  # posterior sampling
  post = coda.samples(jags,
                      jags_params,
                      n.iter = 2500,
                      thin = 5)

  # posterior summaries
  param_summ = post_summ(post,
                         jags_params,
                         Rhat = T,
                         ess = T) %>%
    t() %>%
    as_tibble(rownames = "param") %>%
    mutate(cv = abs(sd / mean))

  # save model results
  file_nm = paste0('RT_only_CJS_', names(y_list)[i], '.rda')

  save(jags_data, post, param_summ,
       file = paste0("analysis/CJS_models/", file_nm))

  rm(y, jags_data, jags, post, param_summ, file_nm)
}

```


# Results

We can extract summary statistics from the posteriors, and construct plots of detection probablities, survival probabilities and cummulative survival probabilities. 

```{r load-results}
# compile parameter estimaets for all years
param_summ_all = c("17_18",
                   "18_19",
                   "19_20") %>%
  as.list() %>%
  rlang::set_names() %>%
  map_df(.id = 'season',
         .f = function(x) {
           load(paste0("../CJS_models/RT_only_CJS_", x, '.rda'))
           param_summ %>%
             mutate(site_num = str_extract(param, "[:digit:]+"),
                    site_num = as.integer(site_num)) %>%
             left_join(tibble(site = colnames(jags_data$y)) %>%
                         mutate(site = factor(site, levels = site),
                                site_num = as.integer(site)) %>%
                         arrange(site_num))
         }) %>%
  mutate(site = factor(site,
                       levels = c("LEMTRP", levels(rec_df$site_code), "LLRTP")),
         site = fct_relevel(site, "LLRTP", after = 8)) %>%
  group_by(season) %>%
  filter(site_num != max(site_num)) %>%
  ungroup()

#----------------------------------------------------------
# reset cummulative survival for 2019-20 to start at LLRTP
load("../CJS_models/RT_only_CJS_19_20.rda")
surv_summ = post %>%
  as.matrix(chain = T,
            iter = T) %>%
  as_tibble() %>%
  select(-(`p[1]`:`phi[7]`),
         -(`survship[1]`:`survship[7]`)) %>%
  mutate(`survship[8]` = 1) %>%
  mutate(`survship[9]` = `survship[8]` * `phi[9]`,
         `survship[10]` = `survship[9]` * `phi[10]`,
         `survship[11]` = `survship[10]` * `phi[11]`,
         `survship[12]` = `survship[11]` * `phi[12]`,
         `survship[13]` = `survship[12]` * `phi[13]`,
         `survship[14]` = `survship[13]` * `phi[14]`,
         `survship[15]` = `survship[14]` * `phi[15]`,
         `survship[16]` = `survship[15]` * `phi[16]`,
         `survship[17]` = `survship[16]` * `phi[17]`) %>%
  select(starts_with('surv')) %>%
  gather(param, value) %>%
  group_by(param) %>%
  summarise(mean = mean(value),
         sd = sd(value),
         `50%` = quantile(value, 0.5),
         `2.5%` = quantile(value, 0.025),
         `97.5%` = quantile(value, 0.975),
         cv = sd / mean) %>%
  ungroup() %>%
  mutate(season = '19_20') %>%
  select(season, everything())


```

```{r phi-tab, echo = F}
param_summ_all %>%
  filter(grepl('phi', param)) %>%
  mutate(reach = paste(lag(site), "to", site)) %>%
  filter(site_num > 1) %>%
  select(season, Reach = reach,
         mean, 
         se = sd,
         cv,
         lwrCI = `2.5%`,
         uprCI = `97.5%`) %>%
  kable(caption = 'Estimates of survival between detection points.',
        digits = 3) %>%
  kable_styling()
```


```{r det-tab, echo = F}
param_summ_all %>%
  filter(grepl('^p\\[', param)) %>%
  filter(site_num > 1) %>%
  select(season, site, 
         mean, 
         se = sd,
         cv,
         lwrCI = `2.5%`,
         uprCI = `97.5%`) %>%
  kable(caption = 'Estimates of detection probability.',
        digits = 3) %>%
  kable_styling()
```


```{r surv-tab, echo = F}
param_summ_all %>%
  filter(grepl('surv', param)) %>%
  select(season, site, 
         mean, 
         se = sd,
         cv,
         lwrCI = `2.5%`,
         uprCI = `97.5%`) %>%
  kable(caption = 'Estimates of cummulative survival up to detection points.',
        digits = 3) %>%
  kable_styling()
```

We also present these results in Figures \@ref(fig:phi-fig), \@ref(fig:det-fig), \@ref(fig:surv-fig) and \@ref(fig:surv2-fig).

```{r make-figures, echo = F}
# plots of estimates
dodge_width = 0.3

det_p = param_summ_all %>%
  filter(grepl('^p\\[', param)) %>%
  ggplot(aes(x = site,
             y = mean,
             color = season)) +
  scale_color_brewer(palette = "Set1",
                     name = "Season") +
  geom_errorbar(aes(ymin = `2.5%`,
                    ymax = `97.5%`),
                width = 0,
                position = position_dodge(width = dodge_width)) +
  geom_point(position = position_dodge(width = dodge_width)) +
  labs(x = 'Site',
       y = 'Detection Probability')

phi_p = param_summ_all %>%
  filter(grepl('phi', param)) %>%
  ggplot(aes(x = site,
             y = mean,
             color = season)) +
  scale_color_brewer(palette = "Set1",
                     name = "Season") +
  geom_errorbar(aes(ymin = `2.5%`,
                    ymax = `97.5%`),
                width = 0,
                position = position_dodge(width = dodge_width)) +
  geom_point(position = position_dodge(width = dodge_width)) +
  labs(x = 'Site',
       y = 'Survival From Previous Site')

surv_p = param_summ_all %>%
  filter(grepl('surv', param)) %>%
  ggplot(aes(x = site,
             y = mean,
             color = season)) +
  scale_color_brewer(palette = "Set1",
                     name = "Season") +
  geom_errorbar(aes(ymin = `2.5%`,
                    ymax = `97.5%`),
                width = 0,
                position = position_dodge(width = dodge_width)) +
  geom_point(position = position_dodge(width = dodge_width)) +
  labs(x = 'Site',
       y = 'Cummulative Survival')

# re-do survival plot focusing on downstream of LLRTP each year
surv_p2 = param_summ_all %>%
  anti_join(surv_summ %>%
              select(season, param)) %>%
  bind_rows(surv_summ %>%
              left_join(param_summ_all %>%
                          select(-(mean:`97.5%`),
                                 -cv))) %>%
  filter(grepl('surv', param)) %>%
  filter(!(season == '19_20' & site_num < 9)) %>%
  ggplot(aes(x = site,
             y = mean,
             color = season)) +
  scale_color_brewer(palette = "Set1",
                     name = "Season") +
  geom_errorbar(aes(ymin = `2.5%`,
                    ymax = `97.5%`),
                width = 0,
                position = position_dodge(width = dodge_width)) +
  geom_point(position = position_dodge(width = dodge_width)) +
  labs(x = 'Site',
       y = 'Cummulative Survival')

```

```{r phi-fig, echo = F, fig.cap = "Estimates, with 95% confidence intervals, of survival between detection points"}
phi_p
```

```{r det-fig, echo = F, fig.cap = "Estimates, with 95% confidence intervals, of detection probabilities."}
det_p
```

```{r surv-fig, echo = F, fig.cap = "Estimates, with 95% confidence intervals, of cummulative survival across reaches."}
surv_p
```

```{r surv2-fig, echo = F, fig.cap = "Estimates, with 95% confidence intervals, of cummulative survival across reaches, resetting survival in 2019-20 to start at Lower Lemhi screw trap."}
surv_p2
```


# Discussion

Some of the caveats to keep in mind with CJS models include:

* Estimates of $\phi$ are really estimates of *apparent* survival. $1 - \phi_i$ represents the chances that an animal either died in reach $i - 1$, or left the population (became unavailable for detection). The latter is analogous to a fish that hunkers down in a reach to ride out the winter. It may be surviving very well within that reach, or it may have died there, there is no way to tell from only the radio tag observations. 
