---
title: "Radio Tag-Only CJS Models"
author:
  - Kevin See:
      email: kevin.see@merck.com
      institute: [biomark]
      correspondence: true
  - Mike Ackerman:
      email: mike.ackerman@merck.com
      institute: [biomark]
  - Nick Porter:
      email: nicholas.porter@merck.com
      institute: [biomark]
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::word_document2:
    fig_height: 6
    fig_width: 7
    toc: false
    always_allow_html: true
    pandoc_args:
    - --lua-filter=templates/scholarly-metadata.lua
    - --lua-filter=templates/author-info-blocks.lua
  bookdown::html_document2:
    fig_height: 6
    fig_width: 7
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    theme: flatly
    pandoc_args:
    - --lua-filter=templates/scholarly-metadata.lua
    - --lua-filter=templates/author-info-blocks.lua
vignette: >
  %\VignetteIndexEntry{Fitting a CJS Model}
  %\VignetteEncoding{UTF-8}
  <!-- %\VignetteEngine{knitr::rmarkdown} -->
  %\VignetteEngine{bookdown::bookdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = F,
  message = F,
  echo = F,
  comment = "#>",
  fig.path = "figures/",
  dpi = 300
)
```

```{r setup, message=F}
# setwd('analysis/writeups')
library(telemetyr)
library(tidyverse)
library(readxl)
library(lubridate)
library(magrittr)
library(janitor)
library(rjags)
library(postpack)
# for formatting in this vignette
library(kableExtra)

# theme_set(theme_bw())
theme_set(theme_classic())

# options for table formating with "kable()"
options(knitr.kable.NA = '-')
options(knitr.table.format = "pandoc")

```

```{r run-model, eval = F}
# this shows how the models were run, but ultimately the results are just read in below
#-------------------------
# set NAS prefix, depending on operating system
#-------------------------
if(.Platform$OS.type != 'unix') {
  nas_prefix = "S:"
}
if(.Platform$OS.type == 'unix') {
  nas_prefix = "~/../../Volumes/ABS"
}


#-------------------------
# read in RT site metadata
#-------------------------
rec_meta = read_excel(paste0(nas_prefix, '/data/telemetry/lemhi/site_metadata/rt_site_metadata.xlsx'))

# which RT receivers were used each year?
# rec_df = rec_meta %>%
#   filter(site_type == 'rt_fixed') %>%
#   filter(as.integer(str_sub(rt_rkm, end = 3)) <= 274) %>%
#   mutate_at(vars(site_code),
#             list(~ factor(., levels = unique(.)))) %>%
#   gather(season, use, starts_with("use")) %>%
#   mutate(season = str_remove(season, "use")) %>%
#   filter(use) %>%
#   select(-use) %>%
#   select(season, everything())
#
#
rec_df = rec_meta %>%
  filter(site_type %in% c('rt_fixed', 'rst')) %>%
  arrange(desc(rt_rkm)) %>%
  select(site = site_code,
         receivers, rt_rkm) %>%
  group_by(site, rt_rkm) %>%
  nest() %>%
  ungroup() %>%
  mutate(receiver = map(data,
                        .f = function(x) {
                          str_split(x, "\\,") %>%
                            extract2(1) %>%
                            str_trim()
                        })) %>%
  select(-data) %>%
  unnest(cols = receiver) %>%
  mutate(receiver = if_else(grepl('NA', receiver),
                            NA_character_,
                            receiver)) %>%
  mutate_at(vars(site, receiver),
            list(~ factor(., levels = unique(.))))


# drop a few upstream sites for these analyses
rec_df %<>%
  filter(!site %in% c('PAHTRP', 'DG', 'KP', 'DC', 'HYDTRP')) %>%
  mutate_at(vars(site, receiver),
            list(fct_drop)) %>%
  mutate(site_num = as.integer(site))


#-------------------------
# write Bayesian CJS model
#-------------------------
model_file = "../CJS_models/CJS_model.txt"
write_bayes_cjs(model_file)

#-------------------------
# load capture histories
#-------------------------
rt_cjs = tibble(season = c('17_18',
                           '18_19',
                           '19_20')) %>%
  mutate(cap_hist_list = map(season,
                             .f = function(yr) {
                               load(paste0(nas_prefix, "/Nick/telemetry/raw/cap_hist_", yr, ".rda"))
                               return(cap_hist_list)
                             }),
         cap_hist_wide = map(cap_hist_list,
                             .f = "ch_wide"),
         tag_meta = map(cap_hist_list,
                        .f = "tag_df"),
         jags_data = map2(cap_hist_wide,
                          tag_meta,
                          .f = function(x, y) {
                            prep_jags_cjs(x, y)
                          }),
         cjs_post = map(jags_data,
                        .f = function(x) {
                          run_jags_cjs(file_path = model_file,
                                       jags_data = x)
                        }),
         param_summ = map2(cjs_post,
                              jags_data,
                             .f = function(x, j_data) {
                               summarise_jags_cjs(x,
                                                  Rhat = T,
                                                  ess = T) %>%
                                 left_join(tibble(site = colnames(j_data$y)) %>%
                                             mutate(site = factor(site, levels = site),
                                                    site_num = as.integer(site))) %>%
                                 select(param_grp, site_num,
                                        site,
                                        param,
                                        everything())
                             }))

param_summ_all = rt_cjs %>%
  select(season, param_summ) %>%
  unnest(cols = param_summ) %>%
  # ignore parameter estimates for last site each season
  group_by(season) %>%
  filter(site_num < max(site_num)) %>%
  ungroup() %>%
  mutate(site = factor(site,
                       levels = levels(rec_df$site))) %>%
  arrange(season, param_grp, site)

# save the results
save(rec_df, rt_cjs, param_summ_all,
     file = '../CJS_models/RT_only_CJS_all.rda')


```


```{r load-data}
load('../CJS_models/RT_only_CJS_all.rda')
```


# Introduction

We are interested primarily in estimating survival of fish from release (e.g. at the Lower Lemhi screw trap) between each RT site using radio tag detections only, with a Cormack Jolly-Seber (CJS) model.

# Methods

## Data

The data is radio tag (RT) detections, which were downloaded from various fixed sites over the last three years. We compressed these detections and summarized capture histories, eliminating any detections that indicated upstream movement. 

In 2017-18 and 2018-19, we focused on tags that were released from the lower Lemhi rotary screw trap (code LLRTP). In 2019-20, we also included tags released from the upper Lemhi rotary screw trap (code LEMTRP). We did remove any tags released at the lower Lemhi trap that were detected upstream in the Lemhi River. There were `r rt_cjs$cap_hist_wide[[3]] %>% filter(LLRTP == 1 & (BC == 1 | TC == 1 | EC == 1 | SS == 1 | EU == 1 | LF == 1)) %>% nrow` tags with a detection history like this.

## Model

We used a Cormack Jolly-Seber model to estimate survival between reaches and detection probability at each detection point. The detection probability at the last site and the survival to that site are confounded in the model, and cannot be estimated separately, and so results from the final reach each year are excluded.

# Results

We can extract summary statistics from the posteriors, and construct plots of detection probabilities, survival probabilities and cumulative survival probabilities. 

```{r load-results}
#----------------------------------------------------------
# reset cummulative survival for 2019-20 to start at LLRTP
# redo posterior samples
surv_llrtp_post = rt_cjs %>%
  filter(season == '19_20') %>%
  pull(cjs_post) %>%
  extract2(1) %>%
  as.matrix(chain = T,
            iter = T) %>%
  as_tibble() %>%
  select(CHAIN, ITER, starts_with("phi")) %>%
  pivot_longer(-(CHAIN:ITER),
               names_to = 'param',
               values_to = 'value') %>%
  mutate(param_grp = str_extract(param, "[:alpha:]+"),
         site_num = str_extract(param, "[:digit:]+"),
         site_num = as.integer(site_num)) %>%
  filter(site_num >= 8) %>%
  mutate_at(vars(value),
            list(~ if_else(site_num == 8,
                           1, .))) %>%
  group_by(CHAIN, ITER) %>%
  mutate(survship = cumprod(value)) %>%
  ungroup() %>%
  mutate(param = str_replace(param, "phi", "survship")) %>%
  select(CHAIN, ITER, param, survship) %>%
  pivot_wider(names_from = "param",
              values_from = "survship")

# summarise survival parameters for 19_20 from LLRTP downstream
surv_summ = rt_cjs %>%
  filter(season == '19_20') %>%
  pull(cjs_post) %>%
  extract2(1) %>%
  as.matrix(chain = T,
            iter = T) %>%
  as_tibble() %>%
  select(-(`survship[8]`:`survship[18]`)) %>%
  left_join(surv_llrtp_post) %>%
  split(list(.$CHAIN)) %>%
  map(.f = as.mcmc) %>%
  as.mcmc.list() %>%
  summarise_jags_cjs(Rhat = T,
                     ess = T) %>%
  mutate(season = '19_20') %>%
  select(season, everything()) %>%
  filter(site_num < max(site_num)) %>%
  left_join(param_summ_all %>%
              select(season, param, site_num, site)) %>%
  select(one_of(names(param_summ_all))) %>%
  arrange(season, param_grp, site_num)

```

```{r phi-tab, echo = F}
param_summ_all %>%
  filter(param_grp == "phi") %>%
  mutate(reach = paste(lag(site), "to", site)) %>%
  filter(site_num > 1) %>%
  select(season, Reach = reach,
         mean, 
         se = sd,
         cv,
         lwrCI = `2.5%`,
         uprCI = `97.5%`) %>%
  kable(caption = 'Estimates of survival between detection points.',
        digits = 3) %>%
  kable_styling()
```


```{r det-tab, echo = F}
param_summ_all %>%
  filter(param_grp == "p") %>%
  filter(site_num > 1) %>%
  select(season, site, 
         mean, 
         se = sd,
         cv,
         lwrCI = `2.5%`,
         uprCI = `97.5%`) %>%
  kable(caption = 'Estimates of detection probability.',
        digits = 3) %>%
  kable_styling()
```


```{r surv-tab, echo = F}
param_summ_all %>%
  filter(grepl('surv', param)) %>%
  select(season, site, 
         mean, 
         se = sd,
         cv,
         lwrCI = `2.5%`,
         uprCI = `97.5%`) %>%
  kable(caption = 'Estimates of cumulative transition up to detection points.',
        digits = 3) %>%
  kable_styling()
```

We also present these results in Figures \@ref(fig:phi-fig), \@ref(fig:det-fig), \@ref(fig:surv-fig) and \@ref(fig:surv2-fig).

```{r make-figures, echo = F}
# plots of estimates
dodge_width = 0.3

det_p = param_summ_all %>%
  filter(param_grp == "p") %>%
  ggplot(aes(x = site,
             y = mean,
             color = season)) +
  scale_color_brewer(palette = "Set1",
                     name = "Season") +
  geom_errorbar(aes(ymin = `2.5%`,
                    ymax = `97.5%`),
                width = 0,
                position = position_dodge(width = dodge_width)) +
  geom_point(position = position_dodge(width = dodge_width)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = 'Site',
       y = 'Detection Probability')

phi_p = param_summ_all %>%
  filter(grepl('phi', param)) %>%
  ggplot(aes(x = site,
             y = mean,
             color = season)) +
  scale_color_brewer(palette = "Set1",
                     name = "Season") +
  geom_errorbar(aes(ymin = `2.5%`,
                    ymax = `97.5%`),
                width = 0,
                position = position_dodge(width = dodge_width)) +
  geom_point(position = position_dodge(width = dodge_width)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = 'Site',
       y = 'Transition Probability From Previous Site')

surv_p = param_summ_all %>%
  filter(grepl('surv', param)) %>%
  ggplot(aes(x = site,
             y = mean,
             color = season)) +
  scale_color_brewer(palette = "Set1",
                     name = "Season") +
  geom_errorbar(aes(ymin = `2.5%`,
                    ymax = `97.5%`),
                width = 0,
                position = position_dodge(width = dodge_width)) +
  geom_point(position = position_dodge(width = dodge_width)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = 'Site',
       y = 'Cumulative Transition Probability')

# re-do survival plot focusing on downstream of LLRTP each year
surv_p2 = param_summ_all %>%
  anti_join(surv_summ %>%
              select(season, param)) %>%
  bind_rows(surv_summ) %>%
  filter(grepl('surv', param)) %>%
  filter(!(season == '19_20' & site_num < 8)) %>%
  ggplot(aes(x = site,
             y = mean,
             color = season)) +
  scale_color_brewer(palette = "Set1",
                     name = "Season") +
  geom_errorbar(aes(ymin = `2.5%`,
                    ymax = `97.5%`),
                width = 0,
                position = position_dodge(width = dodge_width)) +
  geom_point(position = position_dodge(width = dodge_width)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = 'Site',
       y = 'Cumulative Transition Probability')

```

```{r phi-fig, echo = F, fig.cap = "Estimates, with 95% confidence intervals, of transition probabilities between detection points"}
phi_p
```

```{r det-fig, echo = F, fig.cap = "Estimates, with 95% confidence intervals, of detection probabilities."}
det_p
```

```{r surv-fig, echo = F, fig.cap = "Estimates, with 95% confidence intervals, of cumulative transition probability across reaches."}
surv_p
```

```{r surv2-fig, echo = F, fig.cap = "Estimates, with 95% confidence intervals, of cumulative transition probability across reaches, resetting the transitions in 2019-20 to start at Lower Lemhi screw trap."}
surv_p2
```


# Discussion

Some of the caveats to keep in mind with CJS models include:

* Estimates of $\phi$ are really estimates of *apparent* survival, or transition probability (i.e. the chance that a fish survives **AND** moves past the next detection point). $1 - \phi_i$ represents the chances that an animal either died in reach $i - 1$, or left the population (became unavailable for detection). The latter is analogous to a fish that hunkers down in a reach to ride out the winter. It may be surviving very well within that reach, or it may have died there, there is no way to tell from only the radio tag observations. 
